---
title: "Large-sample confidence intervals: proportions"
date: 2021-04-21
---

**aim:** Calculate a **100(1-a)%** approximate confidence interval for a proportion

**reference:** M248 Computer book B, Activity 18

**data:** `accidents.csv`

**description:** Number of accidents suffered by each of 414 machinists over a period of time

## Setup

```{r echo=TRUE, message=FALSE, warning=FALSE}
source("../src/r/load.r")
source("../src/r/confint.r")
library(tidyverse)  # for filtering the data.frame
library(summarytools)
library(ggplot2)
```

```{r}
# import data and assign local vars
accidents <- load_accidents()

# number of machinist who suffered at least one accident
obs <- length(accidents[accidents$Accidents > 0,])
# total number of works in the sample
size <- length(accidents$Accidents)
```

## Describe

### Preview the data

```{r}
head(accidents)
```

### Summarise the data

```{r}
descr(accidents, stat = "common")
```

### Visualise the data

```{r}
ggplot(data = accidents, aes(x = Accidents)) +
  geom_bar()
```

## Results

Calculate a 99% confidence interval for the proportion of workers who suffered at least one accident over the period of observation.

```{r}
zconfint_prop(0.99, obs, size)
```
