---
title: "Large-sample confidence intervals: difference between two proportions"
date: 2021-04-21
---

**aim:** Calculate a **100(1-a)%** approximate confidence interval for the difference between two proportions.

**reference:** M248 Computer book B, Activity 20

**data**: `sewer.csv`

**description**: Occupational risk to sewerage workers of hepatitis A infection though contact with raw sewage.
Relevant fields for this example:

- *Immunity* `0=no immunity, 1=immunity`

- *Children* `1=no children, 2=children`

**note**: A contingency table is constructed to help calculate the relevant proportions.
(And I found this example scenario and data difficult to follow!)

## Setup

```{r echo=TRUE, message=FALSE, warning=FALSE}
source("../src/r/load.r")
source("../src/r/confint.r")
library(summarytools)
library(reshape)
```

```{r}
sewer <- load_sewer()
```

## Describe

### Preview the data

```{r}
head(sewer)
```

## Results

```{r}
# contingency table: children and immunity
(contingency <- table(
  sewer$Children,
  sewer$Immunity,
  dnn = list("children", "immunity")
))
```

```{r}
# marginal totals:  children
rowSums(contingency)
```

Calculate a 95% confidence interval for the difference in the proportions of those infected among sewerage workers with no children and the proportion infected among sewerage workers with children.

```{r}
zconfint_diff_props(
  a = 0.95, # alpha
  x1 = 68,  # has children, has immunity
  n1 = 158, # has children
  x2 = 11,  # no children, has immunity
  n2 = 70   # no children
)
```
