---
title: "Exact confidence intervals for normal means"
date: 2021-04-21
---

**aim:** Calculate a **100(1-a)%** exact confidence interval for a normal mean.

**reference:** M248 Computer book B, Activity 21

**source:** `schoolgirls.csv`

**description:** Height (cm) and weight (kg) of 30 eleven-year-old schoolgirls from Bradford.

## Setup

```{r echo=TRUE, message=FALSE, warning=FALSE}
source("../src/r/load.r")
source("../src/r/confint.r")
library(summarytools)
```

```{r}
# import the sample and assign local var
schoolgirls <- load_schoolgirls()
height <- schoolgirls$Height
weight <- schoolgirls$Weight
```

## About

### Preview the data

```{r}
head(schoolgirls)
```

### Summarise the data

```{r}
descr(schoolgirls, stat = "common")
```

### Visualise the data

```{r}
# histogram of schoolgirl heights
hist(height)
```

## Results

### Check assumption

```{r}
# normal prob plot
qqnorm(height, datax = TRUE); qqline(height, datax = TRUE)
```

### Calculations

Return a 90% confidence interval for the height of schoolgirls.

```{r}
tconfint_mean(
  a = 0.9,
  x = mean(height),
  sd = sd(height),
  n = length(height)
)
```
