---
title: "Exact confidence intervals for the difference between two normal means"
date: 2021-04-21
---

**aim:** Calculate a **100(1-a)%** exact confidence interval for the difference between two normal means.

**reference:** Computer book B, Activity 22

**data:** `skulls.csv`

**description:** Maximum skull breadth (in mm) of 84 skulls of Etruscan males and 70 skulls of modern Italian males.

## Setup

```{r echo=TRUE, message=FALSE, warning=FALSE}
source("../src/r/load.r")
source("../src/r/confint.r")
library(summarytools)
```

```{r}
# import the sample
skulls <- load_skulls()

# assign local vars
etr <- na.omit(skulls$Etruscans)
ita <- na.omit(skulls$Italians)
```

## Describe

### Preview the data

```{r}
head(skulls)
```

### Summary

```{r}
descr(skulls, stat = "common")
```

### Visualise the data

```{r}
boxplot(
  etr,
  ita,
  horizontal=TRUE,
  names=c("Etruscans", "Italians")
)
```

## Results

### Check assumptions

```{r}
# normal prob plot: Etruscan
qqnorm(etr, datax = TRUE); qqline(etr, datax = TRUE)
```

```{r}
# normal prob plot: Italians
qqnorm(ita, datax = TRUE); qqline(ita, datax = TRUE)
```

```{r}
# check for common variance
# note the selection to choose largest variance
if (var(etr) > var(ita)) {
  var(etr)/var(ita)
} else {
  var(etr)/var(ita)
}
```

### Calculations

Return a 95% confidence interval for the difference between the mean skull dimensions.

```{r}
# apologies for the parameter heavy function!
tconfint_diff_means(
  a = 0.95,
  x1 = mean(etr),
  sd1 = sd(etr),
  n1 = length(etr),
  x2 = mean(ita),
  sd2 = sd(ita),
  n2 = length(ita)
)
```
