---
title: "One-sided z-test"
date: 2021-04-22
---

**aim:** Perform a one-sided **z-test** with one sample of data.

**reference:** Computer book B, Exercise 5 *(amended)*

**data:** `practical-test.csv`

**description:** Pass rates for 316 UK driving practical test centres over the period April 2014 to March 2015.

## Setup

```{r echo=TRUE, message=FALSE, warning=FALSE}
source("../src/r/load.r")
library(summarytools)
library(BSDA)
```

```{r}
# import the sample
prac_test <- load_practical_test()

# assign local var to ttoal
rf <- prac_test$Female
```

## Hypotheses

The sample contains the pass rates for 316 UK driving practical test centres over the period April 2014 to March 2015.
During the period April 2013 to March 2014, the national driving practical test pass rate for females was 43.8%.

Let the hypotheses be:

$$
H_{0} : \mu_{F} = 43.8 \%, \> \mu_{F} > 43.8 \%
$$

where $\mu_{2015}$ is the mean female pass rate for the driving practical test nationally across all UK test centres during the period April 2014 to March 2015.

## Method

We will use the function `BSDA::z.test()` to preform the hypothesis test.
See the [documentation](https://www.rdocumentation.org/packages/BSDA/versions/1.2.0/topics/z.test).

## Results

### Preview the data

```{r}
head(rf)
```

### Summarise the data

```{r}
# descriptive stats for female
descr(rf, stats = "common")
```

### Visualise the data

```{r}
hist(rf)
```

### Calculations

```{r}
BSDA::z.test(
  x = rf,
  mu = 43.8,
  sigma.x = sd(rf),
  alternative = "greater")
```

## Discussion

Since $p <$ 0.01, there is strong evidence against the null hypothesis that the mean pass rate of females for the driving practical test nationally from April 2014 to March 2015 is the same as the national pass rate over the same period in the previous year.
We conclude that there is strong evidence that the mean pass rate for females between April 2014 and March 2015 was greater than that for the same period in the previous year.
