---
title: "One-sided z-test for a proportion"
date: 2021-04-22
---

**aim:** Perform a two-sided **z-test** with one sample of data.

**reference:** Computer book B, Activity 26

**data:** **N/A**

## Setup

```{r echo=TRUE, message=FALSE, warning=FALSE}
source("../src/r/load.r")
library(summarytools)
library(BSDA)
```

```{r}
# import the sample
prac_test <- load_practical_test()

# assign local var to total
rt <- prac_test$Total
```

## Hypotheses

The Labour Force Survey (ONS, 2014) of 254 young people aged between 20 and 34 years living in Wales found that 68 were still living at home with their parents.
Previous surveys had found the same proportion to be one in every four young adults.

Let the hypotheses be

$$
H_{0} : p_{W} = 0.25, \> p_{W} \neq 0.25
$$

where the variable $p_{W}$ is the proportion of young adults between 20 and 34 years living with their parents in Wales in 2014.

## Method

We will use the function `stats::prop.test()` to preform the hypothesis test.
See the [documentation](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/prop.test).

## Results

```{r}
(ztest <- prop.test(
  x = 68,    # successes
  n = 254,   # trials
  p = 0.25,  # p0
  correct = FALSE))
```

## Discussion

The $p$-value is 0.514 > 0.1.
There is little to no evidence against the null hypothesis that $p_{W}=0.25$.

## Notes

The test statistic given in the output is `X-squared`.
This is *not* the same as $Z_{P}$, the test statistic used in **M248.**
We can return $Z_{P}$* by calculating `sqrt(X-squared)`

```{r}
sqrt(ztest$statistic)
```

This now matches the value expected by **M248,** but be aware that this will always return a positive value for $Z_{P}$.
Be warned that this will always result in a positive $Z_{P}$.
Just add a negative signed $p < p_{0}$.
