---
title: "Wilcoxon signed rank test (one sample of the differences)"
date: 2021-04-23
---

**aim:** Perform a **Wilcoxon signed rank test** using a sample of the differences.

**reference:** M248 Computer book B, Activity 36

**data:** `movements.csv`

**description:** Differences in the percentage time a foetus moves before a test procedure was performed and the percentage of time it was moving afterwards.

## Setup

```{r echo=TRUE, message=FALSE, warning=FALSE}
source("../src/r/load.r")
library(summarytools)
```

```{r}
# import the data
movements <- load_movements()

# assign local vars
diff <- movements$Diff
```

## Hypotheses

The sample contains data on the differences between the percentage of time a foetus was moving before a test procedure and the percentage of time it was moving afterwards.

Let the hypotheses be

$$
H_{0} : m = 0, \> m \neq 0
$$

where $m$ is the underlying population median difference between pre-test and post-test movement.

## Method

We will use the function `stats::wilcox.test()` to preform the hypothesis test.
See the [documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/wilcox.test.html).

## Results

### Preview the data

```{r}
head(movements)
```

### Summarise the data

```{r}
descr(diff, stats = "common")
```

### Visualise the data

```{r}
boxplot(diff, horizontal = TRUE)
```

### Calculation

```{r}
wilcox.test(
  x = movements$Diff
)
```

## Discussion

Given $p >$ 0.1, there is little to no evidence against the hypothesis that the population median difference between percentages of time a foetus spends moving pre- and post-test are zero.
