---
title: "Wilcoxon signed rank test (two samples)"
date: 2021-04-23
---

**aim:** Perform a **Wilcoxon signed rank test** using two samples of data.

**reference:** M248 Computer book B, Activity 37 (*amended*)

**data:** `lesions.csv`

**description:** Numbers of local lesions appearing on each half of eight leaves, after being rubbed by two preparations of different viruses.

**note:** Methodology differs to how the activity is carried out in the text.
Here we use the two samples of data directly, rather than using a single sample of the calculated differences.

## Setup

```{r echo=TRUE, message=FALSE, warning=FALSE}
source("../src/r/load.r")
library(summarytools)
```

```{r}
# import the data
lesions <- load_lesions()

# assign local vars
a <- lesions$Alesions
b <- lesions$Blesions
```

## Hypotheses

Two virus preparations (labelled **A** and **B**) were soaked into cheesecloth.
The two viruses were then rubbed onto different halves of eight leaves.
The number of local lesions appearing on each half was counted for each of the leaves.

The question arises as to whether, on average, the two virus preparations tend to produce different numbers of lesions.
Suppose that, in particular, the experimenters knew (from the way the preparations were produced) that preparation **B** was very unlikely to lead to more lesions, on average, than preparation **A.**
Hence they wished to carry out a one-sided test with the alternative hypothesis that preparation **A** produces more lesions on average than preparation **B.**

Let the hypotheses be

$$
H_{0} : m_{A} = m_{B}, \> m_{A} > m_{B}
$$

where $m_{A}, \> m_{B}$ are the underlying population medians of the number of lesions found on each half of the leaf after being coated by the two different virus preparations.

## Method

We will use the function `stats::wilcox()` to preform the hypothesis test.
See the [documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/wilcox.test.html).

## Results

### Preview the data

```{r}
head(lesions)
```

### Summarise the data

```{r}
descr(lesions, stats = "common")
```

### Visualise the data

```{r}
boxplot(a, b, horizontal = TRUE, names = c('A', 'B'))
```

### Check assumptions

```{r}
# normal probability plot of the differences
qqnorm(a-b); qqline(a-b)
```

We note that there is a high outlier in the top-right of the plot.

### Calculations

```{r}
wilcox.test(
  x = a,
  y = b,
  alternative = "greater",
  paired = TRUE
)
```

## Discussion

As the $p$-value is 0.015, there is moderate evidence against the null hypothesis.
There is moderate evidence in favour of there being a greater population median number of lesions on the parts treated with virus A compared to B.

We should be cautious though, given the outliers in the normal probability plot, a small sample size of 8 and the number of ties in the sample.
