---
title: "Mann-Whitney Test"
date: 2021-04-23
---

**aim:** Perform a **Mann-Whitney Test** using data with two independent variables.

**reference:** M248 Computer book B, Activity 37 (*amended*)

**data:** `dopamine.csv`

**description:** (See *hypotheses*)

## Setup

```{r echo=TRUE, message=FALSE, warning=FALSE}
source("../src/r/load.r")
library(summarytools)
```

```{r}
dopamine <- load_dopamine()

# assign local var
non_psy <- dopamine$Non.psychotic
psy <- dopamine$Psychotic
```

## Hypotheses

In a study into the causes of schizophrenia, 25 hospitalised patients with schizophrenia were treated with antipsychotic medication, and after a period of time were classified as psychotic or non-psychotic by hospital staff.
A sample of cerebro-spinal fluid was taken from each patient and tested for dopamine $\beta$-hydroxylase enzyme activity.
The measurements are in unit **nmol/(ml)(h)/mg** of protein.

Let the hypotheses be

$$
H_{0} : \ell = 0, \> \ell \neq 0
$$

where $\ell$ is the underlying difference between the locations the populations where the samples were drawn.

## Method

We will use the function `stats::wilcox()` to preform the hypothesis test.
There is no need to `na.omit()` to remove `NaN` results, as this is by the function.
See the [documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/wilcox.test.html).
See also [Unpaired Two-Samples Wilcoxon Test in R](http://www.sthda.com/english/wiki/unpaired-two-samples-wilcoxon-test-in-r) (STDHA).

## Results

### Preview the data

```{r}
head(dopamine)
```

### Summarise the data

```{r}
descr(dopamine, stats = "common", round.digits = 4)
```

### Visualise the data

```{r}
boxplot(
  non_psy, psy,
  horizontal = TRUE,
  names = c('Non Psychotic', 'Psychotic')
)
```

### Results

Return the results of a **Mann-Whitney test.**

```{r}
wilcox.test(
  x = psy,
  y = non_psy,
  alternative = "two.sided",
  correct = FALSE   # adjust for ties
)
```

## Discussion

As the $p$-value is 0.002, there is strong evidence against the null hypothesis.
Looking at the boxplot above, the data suggests that the median $\beta$-hydroxylase enzyme activity in those classed as psychotic is greater than the same measure for non-psychotic patients.
