---
title: "Mann-Whitney Test"
date: 2021-04-23
---

**aim:** Perform a **Mann-Whitney Test** using data with two independent variables.

**reference:** Computer book B, Activity 40

**data:** `osa.csv`

**description:** (See *hypotheses*)

## Setup

```{r echo=TRUE, message=FALSE, warning=FALSE}
source("../src/r/load.r")
library(summarytools)
```

```{r}
osa <- load_osa()

# assign local vars
abs <- osa$Absent
pres <- osa$Present
```


## Hypotheses

A study was carried out of the relationship between customs related to illness and child-rearing practices in 39 non-literate societies.
On the basis of ethnographical reports, each of the societies was given a numerical rating for the degree of something called **oral socialisation anxiety (OSA),** which is a concept derived from psychoanalytic theory relating to child-rearing practice.
(Socialisation refers to the acquiring of social skills.)
Each society was also placed into one of two groups.

The data therefore comprise one sample of **OSA** values for, as it turned out, 23 societies where oral explanations of illness were present, and a second sample of OSA values for 16 societies where oral explanations of illness were absent.

These data were collected to investigate a hypothesis that oral explanations of illness are more likely to be present in societies with high levels of OSA than in societies with low levels of **OSA.**

Let the hypotheses be

$$
H_{0} : \ell_{A} = \ell_{P}, \> \ell_{A} < \ell_{A}
$$

where $\ell_{A} = \ell_{P}$ represent the underlying locations of the populations where the samples were drawn.
(**P**=Present, **A**=Absent.)

## Method

We will use the function `stats::wilcox()` to preform the hypothesis test.
There is no need to `na.omit()` to remove `NaN` results.
See the [documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/wilcox.test.html).
See also [Unpaired Two-Samples Wilcoxon Test in R](http://www.sthda.com/english/wiki/unpaired-two-samples-wilcoxon-test-in-r) (STDHA).

## Results

### Preview the data

```{r}
head(osa)
```

### Summary

```{r}
descr(osa, stats = "common")
```

### Visualise the data

```{r}
boxplot(
  abs, pres,
  horizontal = TRUE,
  names = c('Absent', 'Present'))
```

### Results

Return the results of a **Mann-Whitney test.**

```{r}
wilcox.test(x = abs, y = pres, alternative = "less")
```

## Discussion

The $p$-value is 0.000.
The conclusion is that there is strong evidence against the null hypothesis.
There is strong evidence that societies where oral explanations of illness are present tend to have higher oral socialisation anxiety scores than societies where oral explanations of illness are absent.
