---
title: "Chi-squared goodness-of-fit test"
date: 2021-04-23
---

**aim:** Perform a **chi-squared goodness-of-fit test.**

**reference:** **N/A**

**data:** `uniform_goal.csv`

**description:** (See *hypotheses*)

## Setup

```{r echo=TRUE, message=FALSE, warning=FALSE}
source("../src/r/load.r")
library(summarytools)
library(ggplot2)
library(reshape)
```

```{r}
goals <- load_uniform_goals()

# assign local vars for easier access
group <- goals$group
obs <- goals$obs
```

## Hypotheses

Data has been provided showing the total number of goals scored per ten minute period in an English Premier League football season in normal time.
It has been proposed that a discrete uniform distribution would be a good model for the data.
We will perform a chi-squared goodness-of-fit test to confirm the proposal.

## Method

We will use the function `stats::chisq.test()` to preform the hypothesis test.
See the [documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/wilcox.test.html).
See also [Basic R Guide for NSC Statistics](https://bookdown.org/dli/rguide/chi-square-test.html) (Li, 2020).

## Results

### Preview the data

```{r}
goals
```

### Summarise the data

```{r}
summarytools::descr(
  goals,
  stats = "common"
)
```

### Visualise the data

```{r}
ggplot(goals, aes(group, obs)) +
  geom_col(fill="cornflowerblue") +
  ylab("Frequency")
```

### Calculations

We need to find the probability of an event (i.e., a goal being scored) in each grouping.
As the proposed distribution is a **discrete uniform,** then each grouping should have the same probability of an event.
There are nine groups, so the **Pr** will be

```{r}
# vector of length 9, each element 1/9
pr <- rep(x = 1/9, length.out = 9)
```

Return the results of a **chi-squared goodness-of-fit test.**

```{r}
chisq.test(x = obs, p = pr)
```

Let us compare the **observed** number of goals with the **expected** number of goals.

```{r}
# add expected number of goals
goals$exp <- sum(obs) * pr
# melt columns in long format
mgoals <-melt(goals, id=c("group"))

ggplot(mgoals, aes(group, value, fill = variable)) +
  geom_col(position="dodge") +
  ylab("Frequency") +
  scale_fill_discrete() +
  theme_classic()
```

## Discussion

The $p$-value is 0.03.
This means there is moderate to strong evidence against the null hypothesis.
The data suggests that goals are not uniformly distributed across the 90 minutes of a football match, with them being least likely in the opening ten minutes and most likely between the 60-70 minutes mark.
