---
title: "M248 June 2019"
date: '2021-04-27'
---

## Updates

**2021-04-28:** Initialisation.

## Setup

Path to script: **ljk233/AutomatingM248/src/r/calculators.r**

```{r setup}
source("../src/r/calculators.r")
source("../src/r/confint.r")
```

## Section 1

### q.01

**E.**
*Probability (HB p.6)*

**P**(Classical) = 60/258 $\simeq$ 0.23.

### q.02

**C.**
*Probability (HB p.6)*

**P**(Manager) = 90/258 $\simeq$ 0.35.

### q.03

**B.**
*Probability (HB p.6)*

**P**(Rock | Manager) = 25/90 $\simeq$ 0.28.

### q.04

**D.**
*Probability (HB p.6)*

**P**(Manager | Rock) = 25/98 $\simeq$ 0.26.

### q.05

**A.**
*Probability (HB p.6)*

**P**(Rock or Classical) = (98+60)/258 $\simeq$ 0.61.

### q.06

**C**
*Cumulative distribution functions of a discrete rv (HB p.7)*

The c.d.f. of $X$ at $x=2$ is given by

$$
F(x=2) = P(X \leq 2) = p(1) + p(2) = 0.25+0.5 = 0.7.
$$

### q.07

**E**
*Mean of a discrete rv (HB p.9)*

The expected value $E(X)$ of a discrete random variable $X$ is defined as

$$
E(X) = \sum_{x} x \> p(x).
$$

Therefore, in this scenario,

$$
E(X) = 1(0.2) + 2(0.5) +3(0.3) = 2.1.
$$

### q.08

**B**
*Normalising constants (Book A, p.111)*

Normalising constant $c$ is given by

$$
\begin{aligned}
  1 &= \int_{1}^{3} cx^{3} \> dx = c \int_{1}^{3} cx^{3} \> dx.
\end{aligned}
$$

Therefore, rearranging gives

$$
\frac{1}{c} = \int_{1}^{3} x^{3} \> dx = \bigg[ \frac{1}{4} x^{4} \bigg]_{1}^{3}
  = \frac{1}{4} ( 81 - 1 ) = 20.
$$

Hence, if $c^{-1}$ = 20, then $c$ = 1/20.

```{r q.08}
f <- function (x) {(1/20)*(x**3)}  # declare the fn
integrate(f, 1, 3)  # integrate over range (1,3)
```


### q.09

**F**
*Probabilities of continuous rv (HB p.7)*

If $F(x) = x(2-x), \> 0 < x < 1$, then $P(X \geq 3/4) = 1 - P(X \leq 3/4) = 1 - F(3/4)$.

Therefore,

$$
1 - F(3/4) = 1-\frac{3}{4} \bigg( 2 - \frac{3}{4} \bigg) = 1 - \frac{15}{16} = \frac{1}{16}.
$$

### q.10

**A**
*Probabilities of continuous rv (HB p.7)*

If $F(x) = x(2-x), \> 0 < x < 1$, then 
$P(0 \leq X \leq 1/2) = P(X \leq 1/2) - P(X \leq 0) = F(3/4) - F(0)$.

Therefore,

$$
F(3/4) - F(0) = \frac{1}{2} \bigg( 2 - \frac{1}{2} \bigg) - 0(2-0) = \frac{3}{4} - 0 = \frac{3}{4}.
$$

### q.11

**A**
*Mean of a linear function of rvs (HB p.9)*

For a random variable $Y=3-2X$, then

$$
E(Y) = aE(X) + b,
$$

where $E(X) = 1$, $a = -2$, and $b=3$.
Therefore,

$$
E(Y) = -2 (1) + 3 = 1.
$$

### q.12

**D**
*Variance a linear function of rvs (HB p.9)*

For a random variable $Y=3-2X$, then

$$
S(Y) = \sqrt{V(Y)} = \sqrt{a^{2} V(X)} = |a| \sqrt{V(X)},
$$

where $V(X) = 9$ and $a = -2$.
Therefore

$$
S(Y) = |-2| \sqrt{9} = 6.
$$

### q.13

**C and F.**
*Shape of a normal distribution (Book B p10)*

**C.** True, position is governed by the mean. $E(X) < E(Y)$, so the p.d.f. of $X$ will be to the left of the p.d.f. of $Y$.

**F.** True, the width and height of the p.d.f. of a normal distribution is governed by it standard deviation. $S(Y) > S(Z)$, so the p.d.f. of $Y$ will be wider and flatter than the p.d.f. of $Z$. 

### q.14

**E**
*Probabilities for any non-standard normal (HB p.12)*

If $X \sim N(2.5, 0.16)$, then the probability $P(X > 2.8)$ is given by

$$
\begin{aligned}
  P(X > 2.8) = 1 - P(X < 2.8) &= 1 - P\bigg(Z < \frac{2.8-2.5}{\sqrt{0.16}} \bigg) \\
    &= 1 - \Phi(0.75) \\
    &\simeq 1 - 0.7734 = 0.2266.
\end{aligned}
$$

```{r q.14}
pnorm(2.8, 2.5, sqrt(0.16), lower.tail = FALSE)
```

### q.15

**E**
*Difference between two independent normal rvs (HB p.11)*

If $X \sim N(3,6)$, $Y \sim N(1,5)$, then $U = X - Y$ has the distribution:

$$
E(U) = E(X-Y) = E(X) - E(Y) = 3-1 = 2,
$$

and

$$
V(U) = V(X-Y) = V(X) + V(Y) = 6+5 = 11.
$$

So $U = X - Y \sim N(2, 11)$.

### q.16

**A**
*Quantiles of any non-standard normal (HB p.12)*

The weight of newborn babies (in kg) that is only exceeded by 2% of the newborn babies corresponds to the 0.98-quantile of the distribution of $X \sim N(3.4, 0.5^{2})$, such that

$$
x = \sigma q_{0.98} + \mu,
$$

where $q_{0.98} = 2.054$ is the 0.98-quantile of the standard normal distribution.

Therefore,

$$
x = 0.5(2.036)+3.4 = 4.418 \approx 4.43.
$$

```{r q.16}
qnorm(0.98, 3.4, 0.5)
```

### q.17

**B**
*Sampling distribution of the mean (HB p.12) and Probabilities for any non-standard normal (HB p.12)*

If $X \sim N(502, 5^{2})$, then $\overline{X}_{40}$ will be distributed

$$
E(\overline{X}_{40}) = \mu = 502,
$$

and

$$
V(\overline{X}_{40}) = \frac{\sigma^{2}}{n} = \frac{5^{2}}{40} = 0.625.
$$

Therefore, given $\overline{X}_{40} \sim N(502, 0.625)$, then the probability $P(\overline{X}_{40} < 500)$ will be given by

$$
\begin{aligned}
  P(X < 500) &= 1 - P \bigg( Z < \frac{500-502}{\sqrt{0.625}} \bigg) \\
    &\simeq \Phi(-2.53) \\
    &= 1 - \Phi(2.53) \\
    &= 1 - 0.9943 = 0.0057.
\end{aligned}
$$

```{r q.17}
pnorm(500, 502, sqrt(0.625))
```

### q.18

**C**
*Unbiased estimators (HB p.13)*
If $E(X_{1}) = \theta$, $E(X_{2}) = 3 \theta$, and $E(X_{3}) = 5 \theta$, then an unbiased estimator for $/theta$ is such that

$$
E(\widehat \theta) = \theta.
$$

Looking at the constant and the values of $X_{i}$, let us start with option C,

$$
\begin{aligned}
  E(\widehat \theta) &= E\bigg\{ \frac{1}{11}( 3 X_{1} + X_{2} + X_{3}) \bigg\} \\
    &= \frac{1}{11} E\bigg\{ ( 3 X_{1} + X_{2} + X_{3}) \bigg\} \\
    &= \frac{1}{11} \{ E (3 X_{1}) + E (X_{2}) + E (X_{3}) \} \\
    &= \frac{1}{11} \{ 3 E (X_{1}) + E (X_{2}) + E (X_{3}) \} \\
    &= \frac{1}{11} \{ 3 \theta + 3 \theta + 5 \theta \} \\
    &= \frac{1}{11} \{ 11 \theta \} = \theta.
\end{aligned}
$$

Therefore the correct option is **C**.

### q.19

**C and D.**
*Properties of likelihoods and maximum likelihood estimators (HB p.13,14)*

### q.20

**C**
*Approximate large-sample intervals when the estimators are transformed (HB p.14)*

The transformation $X = \frac{5}{9} (Y-32)$ is both linear and increasing.
When $Y=51.5$,

$$X = \frac{5}{9} (51.5-32) \simeq 10.83,$$

and when $Y=64.6$,

$$X = \frac{5}{9} (64.6-32) \simeq 18.11.$$

Therefore, a new 95% confidence interval in degrees Celsius is approximately $(10.83, 18.11)$.

```{r q.20}
f <- function(x) {5/9*(x-32)}; c(f(51.5), f(64.6))
```

### q.21

**B.**
*Chi-squared goodness-of-fit test (HB p.21)*

The degrees of freedom of a **chi-squared** goodness-of-fit test is $k-p-1$, where

- $k=7$ is the number of categories
- $p=1$ is the number of estimated parameters

Therefore the degrees of freedom is $7-1-1=5$.

### q.22

**E.**
*Interpreting $p$-values (HB p.18)*

There is little or no evidence against the Poisson model fitting these data. 

### q.23

**
*Calculating $S_{xx}$, $S_{yy}$, and $S_{xy}$ (HB p.22)*

Value of $S_{xy}$ is given by

$$
S_{xy} = \sum x_{i} \> y_{i} - \frac{\sum x_{i} \sum y_{i}}{n}.
$$

We know that $n=60$, $\sum x_{i} \> y_{i} = 5954$, $\sum x_{i} = 154$, and $\sum y_{i} = 912$.
Therefore,

$$
S_{xy} = 5954 - \frac{(154)(912)}{60} = 3613.2.
$$

```{r q.23}
calc_slr_std(sum_x = 154, sum_y = 912, sum_sq = 5954, n = 60)
```

### q.24

**C.**
*Calculating the least squares estimate of the intercept $\widehat \alpha$ (HB p.22)*

The value of the least squares estimate of the intercept $\widehat \alpha$ is given by

$$
\widehat \alpha = \overline y - \widehat \beta \overline x = 4.5 - \frac{266}{175} (2.9) = 0.092.
$$

### q.25

**F.**
*Calculating an unbiased estimate s^{2} of \sigma^{2} (HB p.22)*

The value of an unbiased estimate s^{2} of \sigma^{2} is given by

$$
s^{2} = \frac{\sum(y_{i} - \overline{y}^{2})}{n-2} = \frac{84}{18} = 4.666 \ldots
$$

### q.26

**C.**

For a $t$-test to test the hypothesis $H_{0} : \beta = 0,\>H_{1} : \beta \neq 0$ the number of degrees of freedom is $n-2=18$.

### q.27

**A and D.**
*Transforming skew, positive data (HB p.32)*

The data are right-skew, so go down the ladder of powers to make them more normally distributed, so $x^{-1}$ and $\log x$.

### q.28

**E.**
*Transformations and linear regression (HB p.23)*

The residual plot suggests the relationship is non-linear.
Transform the explanatory variable to linearise the relationship.

### q.29

**A.**
*Simple linear regression (HB p.21)*

The predicted value for $y$ given $x=4$ is

$$
y = (1.33 + 4.11 x)^{2} = (1.33 + 4.11 (4))^{2} \simeq 315.8.
$$

### q.30

**F.**
*Elements of a statistical report (HB p.24)*

### q.31

**B.**
*Elements of a statistical report (HB p.24)*

### q.32

**D.**
*Elements of a statistical report (HB p.24)*

## Section 2

### q.33

*Binomial distribution (HB p.8)*

Let $X$ be a discrete random variable that movels whether a patient reacts unfavourably to a certain treatment.
Then $X$ is modelled by the binomial distribution, $X \sim B(n,p)$, where $n=12$ is the sample size and $p=0.03$ is the probability that a patient reacts unfavourably to the treatment, so $X \sim B(12,0.03)$.

The probability $P(X<2) = p(0) + p(1)$ is given by

$$
\begin{aligned}
p(0) + p(1) &=
  \binom{12}{0} 0.03^{0} \> 0.97^{12} + \binom{12}{1} 0.03 \> 0.97^{11} \\
  &\simeq 0.69384 + 0.25750 \\
  &\simeq 0.95135.
\end{aligned}
$$

Therefore, the approximate probability that fewer than patients will react unfavourably to the treatment is 0.951, rounded to 3dp.

```{r q.33}
pbinom(1, 12, 0.03)
```

### q.34

#### (a)

*Cumulative distribution function of a continuous rv (HB p.7)*

The probability that a customer takes longer than 5 minutes to be served is given by $P(t>5) = 1 - F(5)$.
Given the c.d.f. $F(t) = 1 - 4/t^{2}, \> t \geq 2$, then

$$
1 - F(5) = 1 - \bigg( 1 - \frac{4}{5^{2}} \bigg) = \frac{4}{25}.
$$

Therefore, the probability that a customer takes longer than 5 minutes to be served is $\frac{4}{25} = 0.16$.

#### (b)

*Population quantiles of a continuous rv (HB p.11)*

The $\alpha$-quantile of a continuous rv $X$ with c.d.f. $F(X)$ is defined as $F(x) = \alpha$.
If $F(x) = 1 - 4/t^{2}, \> x \in (5,10)$, and the *median*, $m$, is when $\alpha = 1/2$, then

$$
\begin{aligned}
  F(x) = \alpha \to 1 - \frac{4}{m^{2}} &= \frac{1}{2} \\
    \frac{1}{2} &= \frac{4}{m^{2}} \\
    m^{2} &= 8 \\
    m &= \sqrt{8} \\
      &\simeq 2.828.
\end{aligned}
$$

Therefore, the median time taken to serve a customer is approximately 2.83 minutes, rounded to 3dp.

#### (c)

*No reference*

In this scenario, the mean is greater than the median.
This is indicative of data that are right-skew.

## q.35

*Cumulative distribution function of a continuous rv (HB p.7)*

If continuous random variable $X$ has p.d.f. $f(x) = $, then the c.d.f. $F(x)$ will be

$$
$F(x) =
$$

If $f(x) = \frac{1}{12} (5-x)$, where $x \in (0,4)$, then the $F(x)$ will be

$$
\begin{aligned}
F(x) = \int_{0}^{x} \frac{1}{12} (5-y) \> dy
  &= \frac{1}{12} \int_{0}^{x} 5-y \> dy \\
  &= \frac{1}{12} \bigg[ 5y - \frac{1}{2} y^{2} \bigg]_{0}^{x} \\
  &= \frac{1}{12} \bigg( 5x - \frac{1}{2} x^{2} - (0-0) \bigg) \\
  &= \frac{1}{12} \bigg( \frac{1}{2} \bigg) (10x - x^{2}) \\
  &= \frac{1}{24} (10x - x^{2}).
\end{aligned}
$$

## q.36

### (a)

*Geometric distribution (HB p.8) and Bernoulli process (HB p.10)*

Let $S$ be a discrete rv that models the number of customers who arrive at the sports centre in a typical hour who uses the swimming pool, up to and including the first.
Then $S \sim G(p)$, where $p = 0.2$ is the proportion of customers who use the swimming pool.

The probability $P(S=8) = p(8)$, is given by

$$
\begin{aligned}
  p(8) = (1-p)^{7} \> p
    = 0.8^{7} \> 0.2
    = 0.0419 \ldots
    \simeq 0.042.
\end{aligned}
$$

There, the probability the first customer who wants a swim is the 8th customer to arrive in that hour is approximately 0.042, rounded to 3dp.

```{r q.36.a}
dgeom(7, 0.2)
```

### (b)

*Poisson distribution (HB p.8)*

Let $C$ be a discrete rv that models the number of customers who arrive at the sports centre in a given hour.
Then $C \sim \text{Poisson}(\mu)$, where $\mu = 20$ is the number of customers who arrive at the sports centre in a given hour.

The probability 12 customers will arrive at the sports centre, $P(C=20) = p(20)$ is given by

$$
\begin{aligned}
  p(12) = e^{-20} \bigg( \frac{20^{12}}{12!} \bigg) = 0.017625 \ldots
\end{aligned}
$$

Therefore, the probability the 12 customers will arrive at the sports centre in a given hour is approximately 0.0176, rounded to 4dp.

```{r q.36.b}
dpois(12, 20)
```

### (c)

*Binomial distribution (HB p.8)*

Let $Q$ be a discrete rv that models the number of customers who arrive at the sports centre to use the gym in a given hour.
Then $Q \sim B(n,p)$, where $n$ is the number of customers who arrive at the sports centre in the hour and $p=0.15$ is the probability that the customer uses the gym.

If the 15 customers were to arrive in a given hour, then the probability 3 of them will use the gym, $P(Q=3)=p(3)$, is given by

$$
p(3) = \binom{15}{3} 0.15^{3} \> 0.85^{12} = 0.21843 \ldots \simeq 0.218
$$

Therefore, the probability 3 customers will use the gym after arriving at the sports centre is approximately 0.218, rounded to 3dp.

```{r q.36.c}
dbinom(3, 15, 0.15)
```

## q.37

### (a)

*Assumptions for two-sample $t$-intervals (HB p.16)*

The *assumption of an equal variance* is valid if the larger of the sample variances divided by the smaller is less than 3.

We know that $s_{A} = 1.8, \> s_{B} = 1.2$, so

$$
\frac{s_{A}^{2}}{s_{B}^{2}} = \frac{1.8^{2}}{1.2^{2}} = 2.25 < 3.
$$

Therefore, the assumption of equal variances is valid.

```{r q.37.a}
1.8**2; 1.2**2; 1.8**2/1.2**2
```

### (b)

*Exact confidence intervals for the difference between two normal means (HB p.16)*

Given two independent samples from distributions with a common variance, the pooled estimated of the common variance is given by

$$
s_{P}^{2} = \frac{(n_{1}-1) s_{1}^{2} + (n_{2}-1) s_{2}^{2}}{n_{1} + n_{2} - 2}.
$$

We have been given sample standard deviations $s_{A} = 1.8, \> s_{B} = 1.2$, and sample sizes $n_{A} = 8, \> n_{B} = 12$, so

$$
\begin{aligned}
  s_{P}^{2} &= \frac{(8-1) 1.8^{2} + (12-1) 1.2^{2}}{8 + 12 - 2} \\
    &= \frac{(7) 3.24 + (11) 1.44}{18} \\
    &= 2.14.
\end{aligned}
$$

And so the pooled estimate of the common population standard deviation is $\sqrt{s_{P}^{2}} = \sqrt{2.4} \simeq 1.4629$, rounded to 4dp.

```{r q.34.b}
(sp <- calc_pooled_sample_var(v1 = 3.24, n1 = 8, v2 = 1.44, n2 = 12))
sqrt(sp)
```

### (c)

*Exact confidence intervals for the difference between two normal means (HB p.16)*

To calculate a 95% exact confidence interval, we require the 0.975-quantile of the $t(\nu)$ distribution, where $\nu = 18$ degrees of freedom.
Therefore, we require the $0.975$-quantile of $t(18)$, which is $t=2.101$.

```{r q.37.c}
qt(0.975, 18)
```

### (d)

*Exact confidence intervals for the difference between two normal means (HB p.16)*

An exact confidence interval for the difference between the population mean speed increase of serve for male players using method A and male players using method B is given by

$$
\begin{aligned}
  d^{-} &= 8.4 - 7.2 - \bigg\{ 2.101 \sqrt{2.14 \bigg( \frac{1}{8} + \frac{1}{12} \bigg) }\bigg\} \\
    &\simeq -0.203, \\

d^{+} &= 1.2 + 1.1578 \ldots \\
  &\simeq 2.602. \\
\end{aligned}
$$

Hence, an exact 95% confidence interval for the difference between the population mean speed increase of serve for male players using method A and male players using method B is $(-0.203, 2.602)$.

The confidence interval contains zero, so it does not support a claim that there is a difference between the two methods.

```{r}
tconfint_diff_means(
  a = 0.95, x1 = 8.4, sd1 = 1.8, n1 = 8, x2 = 7.2, sd2 = 1.2, n2 = 12)
```

### (e)

*Repeated experiments interpretation of confidence intervals (HB p.14)*

If a large number of samples of size 10 were drawn independently from the population of women coached using method A, and a large number of samples of size 10 were drawn independently from the population of women coached using method B, and the difference, and a 95% confidence interval for the mean difference in the increase in serve speed calculated in each occasion, then approximately 95% of these intervals would contain the true mean difference in the increase in serve speed.

The 90% confidence interval actually observed, (-4.01, 1.12), is just one observation on a random interval, and may or may not contain the population mean.

## q.38

```{r q.38}
(ztest <- prop.test(
    x = 38, n = 50, p = 0.6, alternative = "greater", correct = FALSE))
```

### (a)

*Testing a population proportion (HB p.17)*

Let the hypotheses be

$$
H_{0} : p = 0.6, \>  p_{X} > 0.6
$$

where $p$ is the proportion of 13-year-olds who reach  standard X, in physical education.

### (b)

*Testing a population proportion (HB p.17)*

The sample size is large, so the test statistic $Z_{p_{X}}$ with distribution $N(0, 1)$ is given by

$$
Z_{p} = \frac{\widehat{p} - p_{0}}{\sqrt{\frac{p_{0}(1-p_{0})}{n}}},
$$

where $\widehat{p} = X/n = 38/50 = 0.76,\>p_{0} = 0.6,\>n=50$.
Therefore

$$
Z_{p} = \frac{0.76-0.6}{\sqrt{\frac{0.6(0.4)}{50}}} = \frac{0.16}{0.069 \ldots} \simeq 2.31.
$$

Hence, the observed value test statistic $Z_{p}$ is approximately 2.31, rounded to 2dp.

```{r q.38.b}
sqrt(ztest$statistic)
```


### (c)

*Calculating p-value (HB p.18)*

The $p$-value for this one-sided test of a proportion is

$$
P(Z_{p} > z_{p}) = 1 - P(Z<2.31) = 1-0.9896 = 0.0104.
$$

```{r q.38.c}
ztest$p.value
```

### (d)

*Interpreting $p$-values (HB p.18)*

Given $p$-value $\in (0.01, 0.05]$, there is moderate to strong evidence against the null hypothesis.
We conclude that there is moderate to strong evidence that more than 60% of 13-year-olds reach standard X is physical education.

## q.39

### (a)

*Calculating the test statistic of the Wilcoxon signed rank test (HB p.19)*

The table in the question was copied and extended to included the absolute difference and the rank of the absolute difference.

| Bed partner    | 1 | 2   | 3   | 4 | 5   | 6   | 7 | 8 | 9   | 10  |
|----------------|---|-----|-----|---|-----|-----|---|---|-----|-----|
| Difference     | 7 | 3   | −2  | 0 | 1   | −1  | 4 | 6 | 2   | 3   |
| Abs.difference | 7 | 3   | 2   | 0 | 1   | 1   | 4 | 6 | 2   | 3   |
| Rank           | 9 | 5.5 | 3.5 | * | 1.5 | 1.5 | 7 | 8 | 3.5 | 5.5 |

The Wilcoxon signed rank test statistic $W_{+}$ is the sum of the ranks of the positive differences, so

$$
w_{+} = 9+5.5+1.5+7+8+3.5+5.5 = 40
$$

```{r}
9+5.5+1.5+7+8+3.5+5.5
```

Therefore, te test statistic is $W_{+} = 40$.

### (b)

*Normal approximation to the null distribution of the Wilcoxon test statistic (HB p.20)*

Under the null hypothesis, for a sample of size $9$ (excluding the zero difference), the Wilcoxon signed rank test statistic $W_{+}$ has mean,

$$
\begin{aligned}
  E(W_{+}) = \frac{n(n+1)}{4} = \frac{9(10)}{4} = 22.5,
\end{aligned}
$$

and variance,

$$
\begin{aligned}
  V(W_{+}) = \frac{n(n+1)(2n+1)}{24} = \frac{9(10)(19)}{24} = 71.25.
\end{aligned}
$$

```{r}
# returns (z, Ew, Vw)
calc_norm_approx_wilcoxon(40, 9)
```

### (c)

*Normal approximation to the null distribution of the Wilcoxon test statistic (HB p.20) and ==Missing== (HB p.18)*

In this scenario, the observed test statist is $w_{+} = 40$, so the corresponding observed value of $Z$ is

$$
Z = \frac{w_{+} - E(W_{+})}{\sqrt{V(W_{+})}} = \frac{40-22.5}{\sqrt{71.25}} \simeq 2.07.
$$

### (d)

*Calculating p-values (HB p.18)*

Using the table of probabilities of the standard normal distribution in the handbook

$$
P(Z > 2.07) = 1 - P(Z < 2.07) = 1 - 0.9808 = 0.0192.
$$

This is a two-sided test, so the $p$-value for this test will be $2(0.0192) = 0.0384$.

### (e)

*Interpreting p-values (HB p.18)*

Since $p \in (0.01, 0.05]$ there is moderate to strong evidence against the null hypothesis.
Given the test statistic is greater than zero, there is evidence to suggest that the palatal transplants had a positive effect.

### (f)

*Normal approximation to the null distribution of the Wilcoxon test statistic (HB p.20)*

The normal approximation of the Wilcoxon signed ranked test is adequate for sample size greater than or equal to 16.
The sample size in this experiment is smaller than that ($n=9$ after the removal of zero differences).
This makes the normal approximation here possibly unreliable, so may explain the difference between the approximate and exact $p$-values.
